---
title: "Exercise 7"
author: "Paulo Arevalo"
date: "March 21, 2016"
output: html_document
---

```{r}
if(!require(PEcAn.data.land)){
  install.packages(c("digest","dplR"), repos='http://cran.us.r-project.org')
  devtools::install_github("PecanProject/pecan/utils")
  devtools::install_github("PecanProject/pecan/db")
  devtools::install_github("PecanProject/pecan/modules/data.land")
  require(PEcAn.data.land)
}
require(rjags)
```

```{r, echo=FALSE}
#plots a confidence interval around an x-y plot (e.g. a timeseries)
ciEnvelope <- function(x,ylo,yhi,...){
  polygon(cbind(c(x, rev(x), x[1]), c(ylo, rev(yhi),
                                      ylo[1])), border = NA,...) 
}
```


```{r}
## 1. Read tree data
trees <- read.csv("data/H2012AdultFieldData.csv")

## 2. Read tree ring data
rings <- Read_Tucson("data/TUCSON/")

## 3. merge inventory and tree ring data, extract most recent nyears
combined <- matchInventoryRings(trees,rings,nyears=17)

## take a look at the first few rows of data to see the structure
knitr::kable(combined[1:5,])

## 4. organize data into a list
data <- buildJAGSdata_InventoryRings(combined)

# y = increment (tree x year)
# z = dbh (tree x year)
# make sure to take a look at all the priors!
```

# Fixed effect (Base case)

```{r, fig.height=8}

n.iter = 20000

## this code fuses forest inventory data with tree growth data (tree ring or dendrometer band)
## for the same plots. Code is a rewrite of Clark et al 2007 Ecol Appl into JAGS
TreeDataFusionBase = "
model{

  ### Loop over all individuals
  for(i in 1:ni){
  
  #### Data Model: DBH
  for(t in 1:nt){
  z[i,t] ~ dnorm(x[i,t],tau_dbh)
  }
  
  #### Data Model: growth
  for(t in 2:nt){
  inc[i,t] <- x[i,t]-x[i,t-1]
  y[i,t] ~ dnorm(inc[i,t],tau_inc)
  }
  
  #### Process Model
  for(t in 2:nt){
  Dnew[i,t] <- x[i,t-1] + mu
  x[i,t]~dnorm(Dnew[i,t],tau_add)
  }
  
  x[i,1] ~ dnorm(x_ic,tau_ic)
  }  ## end loop over individuals
  
  #### Priors
  tau_dbh ~ dgamma(a_dbh,r_dbh)
  tau_inc ~ dgamma(a_inc,r_inc)
  tau_add ~ dgamma(a_add,r_add)
  mu ~ dnorm(0.5,0.5)
}"

  ## state variable initial condition
  z0 = t(apply(data$y,1,function(y){-rev(cumsum(rev(y)))})) + data$z[,ncol(data$z)] 
  
  ## JAGS initial conditions
  nchain = 3
  init <- list()
  for(i in 1:nchain){
    y.samp = sample(data$y,length(data$y),replace=TRUE)
    init[[i]] <- list(x = z0,tau_add=runif(1,1,5)/var(diff(y.samp),na.rm=TRUE),
                      tau_dbh=1,tau_inc=500,tau_ind=50,tau_yr=100,ind=rep(0,data$ni),year=rep(0,data$nt))
  }
  
  ## compile JAGS model
  j.model   <- jags.model (file = textConnection(TreeDataFusionBase),
                           data = data,
                           inits = init,
                           n.chains = 3)
  ## burn-in
  jags.out   <- coda.samples (model = j.model,
                              variable.names = c("tau_add","tau_dbh","tau_inc","mu","tau_ind","tau_yr"),
                              n.iter = min(n.iter,2000))
  plot(jags.out)
  
  ## run MCMC
  jags.out   <- coda.samples (model = j.model,
                              variable.names = c("x","tau_add","tau_dbh","tau_inc","mu",
                                                 "tau_ind","tau_yr","ind","year"),
                              n.iter = n.iter)
```

## Diagnostic plots - base case

```{r, fig.height=8}
  #### Diagnostic plots
  
  ### DBH
  layout(matrix(1:8,4,2))
  out <- as.matrix(jags.out)
  x.cols = which(substr(colnames(out),1,1)=="x")   ## which columns are the state variable, x
  ci <- apply(out[,x.cols],2,quantile,c(0.025,0.5,0.975))
  ci.names = parse.MatrixNames(colnames(ci),numeric=TRUE)
  
  smp = c(sample.int(data$ni,3),49)  ## I've rigged the sampling to make sure you see tree 49!
  for(i in smp){
    sel = which(ci.names$row == i)
    plot(data$time,ci[2,sel],type='n',ylim=range(ci[,sel],na.rm=TRUE),ylab="DBH (cm)",main=i)
    ciEnvelope(data$time,ci[1,sel],ci[3,sel],col="lightBlue")
    points(data$time,data$z[i,],pch="+",cex=1.5)
  }
  
  ## growth
  for(i in smp){
    sel = which(ci.names$row == i)
    inc.mcmc = apply(out[,x.cols[sel]],1,diff)
    inc.ci = apply(inc.mcmc,1,quantile,c(0.025,0.5,0.975))*5
    
    plot(data$time[-1],inc.ci[2,],type='n',ylim=range(inc.ci,na.rm=TRUE),ylab="Ring Increment (mm)")
    ciEnvelope(data$time[-1],inc.ci[1,],inc.ci[3,],col="lightBlue")
    points(data$time,data$y[i,]*5,pch="+",cex=1.5,type='b',lty=2)
  }
```

## Mixed effects, mu and precision - base case

```{r}
  ## process model
  vars = (1:ncol(out))[-c(which(substr(colnames(out),1,1)=="x"),grep("tau",colnames(out)),
                          grep("year",colnames(out)),grep("ind",colnames(out)))]
  par(mfrow=c(1,1))
  for(i in vars){
    hist(out[,i],main=colnames(out)[i])
  }
  if(length(vars)>1) pairs(out[,vars])

  ## Standard Deviations
  #layout(matrix(c(1,2,3,3),2,2,byrow=TRUE))
  par(mfrow=c(2,3))
  prec = out[,grep("tau",colnames(out))]
  for(i in 1:ncol(prec)){
    hist(1/sqrt(prec[,i]),main=colnames(prec)[i])
  }
  cor(prec)
  pairs(prec)
```

# Random effects (Extended case)

```{r, fig.height=8}

n.iter = 20000
TreeDataFusionExtended = "
  
model{
  
  ### Loop over all individuals
  for(i in 1:ni){
  
  #### Data Model: DBH
  for(t in 1:nt){
    z[i,t] ~ dnorm(x[i,t],tau_dbh)
  }
  
  #### Data Model: growth
  for(t in 2:nt){
    inc[i,t] <- x[i,t]-x[i,t-1]
    y[i,t] ~ dnorm(inc[i,t],tau_inc)
  }
  
  #### Process Model
  for(t in 2:nt){
    Dnew[i,t] <- x[i,t-1] + mu + ind[i] + year[t]
    x[i,t]~dnorm(Dnew[i,t],tau_add)
  }
  
  ## individual effects
  ind[i] ~ dnorm(0,tau_ind)
  
  ## initial condition
  x[i,1] ~ dnorm(x_ic,tau_ic)
  }  ## end loop over individuals
  
  ## year effects
  for(t in 1:nt){
    year[t] ~ dnorm(0,tau_yr)
  }
  
  
  #### Priors
  tau_dbh ~ dgamma(a_dbh,r_dbh)
  tau_inc ~ dgamma(a_inc,r_inc)
  tau_add ~ dgamma(a_add,r_add)
  tau_ind ~ dgamma(1,0.1)
  tau_yr  ~ dgamma(1,0.1)
  mu ~ dnorm(0.5,0.5)
  
  }"
  
  z0 = t(apply(data$y,1,function(y){-rev(cumsum(rev(y)))})) + data$z[,ncol(data$z)] 
  
  ## JAGS initial conditions
  nchain = 3
  init <- list()
  for(i in 1:nchain){
    y.samp = sample(data$y,length(data$y),replace=TRUE)
    init[[i]] <- list(x = z0,tau_add=runif(1,1,5)/var(diff(y.samp),na.rm=TRUE),
                      tau_dbh=1,tau_inc=500,tau_ind=50,tau_yr=100,ind=rep(0,data$ni),year=rep(0,data$nt))
  }
  
  ## compile JAGS model
  j.model2   <- jags.model (file = textConnection(TreeDataFusionExtended),
                           data = data,
                           inits = init,
                           n.chains = 3)
  ## burn-in
  jags.out2   <- coda.samples (model = j.model2,
                              variable.names = c("tau_add","tau_dbh","tau_inc","mu","tau_ind","tau_yr"),
                              n.iter = min(n.iter,2000))
  plot(jags.out2)
  
  ## run MCMC
  jags.out2   <- coda.samples (model = j.model2,
                              variable.names = c("x","tau_add","tau_dbh","tau_inc","mu",
                                                 "tau_ind","tau_yr","ind","year"),
                              n.iter = n.iter)
```
  
## Diagnostic plots - extended case

```{r, fig.height=8}
  #### Diagnostic plots
  
  ### DBH
  layout(matrix(1:8,4,2))
  out <- as.matrix(jags.out2)
  x.cols = which(substr(colnames(out),1,1)=="x")   ## which columns are the state variable, x
  ci <- apply(out[,x.cols],2,quantile,c(0.025,0.5,0.975))
  ci.names = parse.MatrixNames(colnames(ci),numeric=TRUE)
  
  for(i in smp){
    sel = which(ci.names$row == i)
    plot(data$time,ci[2,sel],type='n',ylim=range(ci[,sel],na.rm=TRUE),ylab="DBH (cm)",main=i)
    ciEnvelope(data$time,ci[1,sel],ci[3,sel],col="lightBlue")
    points(data$time,data$z[i,],pch="+",cex=1.5)
  }
  
  ## growth
  for(i in smp){
    sel = which(ci.names$row == i)
    inc.mcmc = apply(out[,x.cols[sel]],1,diff)
    inc.ci = apply(inc.mcmc,1,quantile,c(0.025,0.5,0.975))*5
    
    plot(data$time[-1],inc.ci[2,],type='n',ylim=range(inc.ci,na.rm=TRUE),ylab="Ring Increment (mm)")
    ciEnvelope(data$time[-1],inc.ci[1,],inc.ci[3,],col="lightBlue")
    points(data$time,data$y[i,]*5,pch="+",cex=1.5,type='b',lty=2)
  }
```

## Mixed effects, mu and precision - extended case

```{r}
  ## process model
  vars = (1:ncol(out))[-c(which(substr(colnames(out),1,1)=="x"),grep("tau",colnames(out)),
                          grep("year",colnames(out)),grep("ind",colnames(out)))]
  par(mfrow=c(1,1))
  for(i in vars){
    hist(out[,i],main=colnames(out)[i])
  }
  if(length(vars)>1) pairs(out[,vars])

  ## Standard Deviations
  #layout(matrix(c(1,2,3,3),2,2,byrow=TRUE))
  par(mfrow=c(2,3))
  prec = out[,grep("tau",colnames(out))]
  for(i in 1:ncol(prec)){
    hist(1/sqrt(prec[,i]),main=colnames(prec)[i])
  }
  cor(prec)
  pairs(prec)
```

## Random effects - extended case

```{r}
  par(mfrow=c(1,1))
  ### YEAR
  year.cols = grep("year",colnames(out))
  if(length(year.cols>0)){
    ci.yr <- apply(out[,year.cols],2,quantile,c(0.025,0.5,0.975))
    plot(data$time,ci.yr[2,],type='n',ylim=range(ci.yr,na.rm=TRUE),main="Year Effect",ylab="cm")
    ciEnvelope(data$time,ci.yr[1,],ci.yr[3,],col="lightBlue")
    lines(data$time,ci.yr[2,],lty=1,lwd=2)
    abline(h=0,lty=2)
  }
  
  ### INDIV
  ind.cols= which(substr(colnames(out),1,3)=="ind")
  if(length(ind.cols)>0){
    boxplot(out[,ind.cols],horizontal=TRUE,outline=FALSE,col=combined$PLOT,main="Individual Effects By Plot",xlab="cm")
    abline(v=0,lty=2)
    ## calculate plot-level means for random effects
    tapply(apply(out[,ind.cols],2,mean),combined$PLOT,mean)
    table(combined$PLOT)
    
    spp = combined$SPP
    boxplot(out[order(spp),ind.cols],horizontal=TRUE,outline=FALSE,col=spp[order(spp)],main="Individual Effects By Species",xlab="cm")
    abline(v=0,lty=2)
    spp.code = levels(spp)[table(spp)>0]
    legend("bottomright",legend=rev(spp.code),col=rev(which(table(spp)>0)),lwd=4)
    ## calculate species-level means for random effects
    tapply(apply(out[,ind.cols],2,mean),combined$SPP,mean)
  }
```

Summary:

The diagnostic plots show how the inclusion of random effects (or any effects that account for some aspect of the data or the process in question, for that matter) can reduce the uncertainty in the prediction. Looking at the individual effects by by plot graph, there is no apparent separation between different species by plot. On the other hand, individual effects by species do seem to display some separation between at least some of them. For this reason, we could used species as a fixed or random effect to account for some of the unexplained variability in the estimates. One way to analyze this relationship would be to calculate the overall mean increase in DBH and plot it against the species to see the levels of separability between them (i.e. a boxplot). We could also use additional covariates like precipitation to determine if some of the unexplained variation is caused by increase water availability, and in this case we would also plot the fixed effect against the data to determine if the supposed relationship holds true, and in which magnitude. 

Proposed code:

```
TreeDataFusionPlus = "
  
model{
  
  ### Loop over all individuals
  for(i in 1:ni){
  
  #### Data Model: DBH
  for(t in 1:nt){
    z[i,t] ~ dnorm(x[i,t],tau_dbh)
  }
  
  #### Data Model: growth
  for(t in 2:nt){
    inc[i,t] <- x[i,t]-x[i,t-1]
    y[i,t] ~ dnorm(inc[i,t],tau_inc)
  }
  
  #### Process Model
  # Species effect, which has an matrix/array with the same number of individuals, allowing indexing it with 
  # i, and dummy variables for the species. The pseudocode makes reference to the need to specify a logical
  # condition in order to bring up the correct distribution for the given species in that iteration
  
  for(t in 2:nt){
      Dnew[i,t] <- x[i,t-1] + mu + ind[i] + year[t] + dummy[i,] * (<condition to select> spp[]) 
      x[i,t]~dnorm(Dnew[i,t],tau_add)
    
  }
  
  ## individual effects
  ind[i] ~ dnorm(0,tau_ind)
  
  ## initial condition
  x[i,1] ~ dnorm(x_ic,tau_ic)
  }  ## end loop over individuals
  
  # INCLUDE species effect, where dist corresponds to any particular distribution that best describes the effect of each species (or genus or functional group if we wanted) in the slope of the DBH growth change over time. 
  
  for(s in 1:ns){
    spp[s] ~ dist(spp_prior) 
  }
  
  ## year effects
  for(t in 1:nt){
    year[t] ~ dnorm(0,tau_yr)
  }
  
  
  #### Priors
  tau_dbh ~ dgamma(a_dbh,r_dbh)
  tau_inc ~ dgamma(a_inc,r_inc)
  tau_add ~ dgamma(a_add,r_add)
  tau_ind ~ dgamma(1,0.1)
  tau_yr  ~ dgamma(1,0.1)
  mu ~ dnorm(0.5,0.5)
  
   # We'd need to specify the distribution that best describes the prior for the species, with its corresponding hyperparameters, based on any knowledge we might have on how the species growth rate may differ.
   
  spp_prior ~ dist(HYPERPARAMETERS)
  
  }"

```
